---
title: "Control_Y1_viability"
output:
  pdf_document: default
  html_document: default
date: "2026-08-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
Control_Y1 <- read.csv("../output/Control_Y1_w_elev.csv")
```

```{r}
control_1 <- Control_Y1 %>%
  mutate (
    viable = Updated_viability_cat == 1 #set only consider category 1 as viable
    )

control_sum_1 <- control_1 %>%
  group_by (Pop, Elev_m) %>%
  summarise (
    trials = n(), #count total number of tested seeds
    events = sum(viable), #count the number of viable seeds
    proportion = events / trials, #calculate the proportion of viable
    SE = sd(viable) / sqrt(n()), #calculate SE
    .groups = "drop"
  )
print(control_sum_1)
```


```{r}
control_glm_1 <- glm( #fit a binomial generalized linear model
  viable ~ Pop, #model the probability that an individual seed is viable as a function of population
  family = binomial, 
  data = control_1
)
print(summary(control_glm_1)) #can also run a population rather than population
```
###Seed viability differed among source populations under laboratory storage, indicating that populations retained different baseline levels of seed viability after storage.

```{r}
anova(control_glm_1, test = "Chisq")
```
###Seed viability differed significantly among source populations under laboratory storage (Analysis of Deviance, χ² = 41.55, df = 10, P < 0.001).

```{r}
control_prediction_1 <- control_sum_1 %>%
  select (Pop)
control_prediction_1$predicted <- predict( #calculate predicted viability for each population
  control_glm_1, #use fitted binomial glm
  newdata = control_prediction_1, #use predicted new population data
  type = "response" #return predicted viability data
)
print(control_prediction_1)
```


```{r}
control_figure_1 <- ggplot( #create a ggplot use population level summary data with x-axis in population ordered by ascending elevation in meter and y-axis by mean proportion of viable
  control_sum_1, aes (x = reorder(Pop, Elev_m), y = proportion) 
) +
  geom_point(size = 3) + #each populations' mean proportion of viable
   geom_errorbar( #add SE bar
    aes(
      ymin= proportion - SE,
      ymax = proportion + SE
    ),
    width = 0.2
  ) +
   geom_line( #add binomial glm trendline by using new prediction data
    data = control_prediction_1,
    aes(x = Pop, y = predicted, group = 1), #plot predicted viability across population
    linewidth = 1, color = "red"
  ) +
  theme_classic () + #use classic ggplot theme
  labs ( #add labels for x-axis and y-axis
    x = "Population (ordered by ascending elevation in meter)",
    y = "Proportion Viable"
  )
print (control_figure_1)
```

