---
title: "CP2_Y1_viability"
output:
  pdf_document: default
  html_document: default
date: "2026-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```


```{r}
library(tidyverse)
library(ggplot2)
#library(ggbreak) 
```


```{r}
CP2_Y1 <- read.csv("../output/CP2_Y1_w_elev.csv") %>%
  filter(!(Pop == "WL1" & Rep == 14)) #filter WL1 rep14, because there is no seeds in WL1 rep14 but we have marked it in the spreadsheet; without filter, the NA will be include in the SE and cause all the reps of WL1 become NA
```


```{r}
rep_sum_1 <- CP2_Y1 %>%
  group_by (Pop, Elev_m, Rep) %>% #group seeds data by population, elevation and replicates
  summarise (
    trials = sum (!is.na(Updated_viability_cat)), #count how many seeds we have tested in each replicate and remove the cell with NA data in the Updated_viability_cat (some rows are empty seed coat or like WL1 rep14 is an empty envelope)
    events = sum (Updated_viability_cat == 1, na.rm = TRUE), #count how many seeds are considered as viable (category 1)
    .groups = "drop"
  ) %>%
  mutate (proportion = events/trials) #calculate viable proportion of each replicate

pop_sum_1 <- rep_sum_1 %>%
  group_by (Pop, Elev_m) %>% #group each replicate data by population and elevation
  summarise(
    n_reps = n(), #count number of replicates of each population
    mean_proportion = mean (proportion, na.rm = TRUE), #calculate mean proportion viable across replicates
    SE = sd(proportion, na.rm = TRUE)/sqrt(n()), #calculate SE of mean viability
    .groups = "drop"
  )
print(pop_sum_1)
```



```{r}
glm_1 <- glm( #fit a binomial generalized linear model
  cbind(events, trials - events) ~ Elev_m, #model viable and non-viable seeds count as a function of population
  family= binomial, #use binomial distribution
  data = rep_sum_1
)
anova(glm_1, test = "Chisq")

prediction_1 <- data.frame( #create a prediction dataset containing one row for each population
  Elev_m = seq(
    min(rep_sum_1$Elev_m),
    max(rep_sum_1$Elev_m),
    length.out = 100
  )
)
prediction_1$predicted <- predict( #calculate predicted viability for each elevation
  glm_1, #use fitted binomial glm
  newdata = prediction_1, #use predicted new elevation data
  type = "response" #return predicted viability data
)

print(summary(glm_1))
```
###Different source populations shows different viability following one year of burial at the common garden
###Seed viability differed significantly among source populations following one year of burial at the common garden (Analysis of Deviance, χ² = 66.62, df = 10, P < 0.001).


glm_1 <- glm( #fit a binomial generalized linear model
  cbind(events, trials - events) ~ Pop, #model viable and non-viable seeds count as a function of population
  family= binomial, #use binomial distribution
  data = rep_sum_1
)
anova(glm_1, test = "Chisq")

prediction_1 <- pop_sum_1 %>% #create a prediction dataset containing one row for each population
  select(Pop)
prediction_1$predicted <- predict( #calculate predicted viability for each elevation
  glm_1, #use fitted binomial glm
  newdata = prediction_1, #use predicted new elevation data
  type = "response" #return predicted viability data
)

print(summary(glm_1))
print(prediction_1)




figure_1 <- ggplot(pop_sum_1, aes (x = reorder(Pop, Elev_m), y = mean_proportion)) + #create a ggplot use population level summary data with x-axis in elevation by meter and y-axis by mean proportion of viable
  geom_point(size = 3) + #each populations' mean proportion of viable
  geom_errorbar( #add SE bar
    aes(
      ymin= mean_proportion - SE,
      ymax = mean_proportion + SE
    ),
    width = .3
  ) +
  geom_line( #add binomial glm trendline by using new prediction data
    data = prediction_1,
    aes(x = Pop, y = predicted, group = 1), #plot predicted viability across elevation
    linewidth = 1, color = "red"
  ) +
  theme_classic() + #use classic ggplot theme
  labs ( #add labels for x-axis and y-axis
    x = "Population (ordered by ascending elevation in meter)",
    y = "Proportion Viable"
  )
print(figure_1)

```{r}
figure_1 <- ggplot(pop_sum_1, aes (x = Elev_m, y = mean_proportion, color = Pop)) + #create a ggplot use population level summary data with x-axis in elevation by meter and y-axis by mean proportion of viable
  geom_point(size = 3) + #each populations' mean proportion of viable
  geom_errorbar( #add SE bar
    aes(
      ymin= mean_proportion - SE,
      ymax = mean_proportion + SE, color = Pop
    ),
    width = 50
  ) +
  #scale_x_break(c(600, 1000)) +
  #scale_x_break(c(1100, 1600)) +
  #scale_x_break(c(1700, 2000)) +
  geom_text (aes (label = Pop), vjust = -5, size = 3, show.legend = FALSE) + #label each point with its population name
  geom_line( #add binomial glm trendline by using new prediction data
    data = prediction_1,
    aes(x = Elev_m, y = predicted, group = 1), #plot predicted viability across elevation
    linewidth = 1, color = "red"
  ) +
  scale_x_continuous(limits = c(350, 2550), n.break = 15) +
  scale_y_continuous(limits = c(.35, .85)) +
  theme_classic() + #use classic ggplot theme
  theme(text=element_text(size=10), axis.text.x = element_text(angle = 45,  hjust = 1)) +
  labs ( #add labels for x-axis and y-axis
    x = "Population (ordered by ascending elevation in meter)",
    y = "Proportion Viable"
  ) 
print(figure_1)
```


to do:
geomjitter can fix the population stack together

```{r}
distance <- read.csv("../output/CP2_Y1_sites_w_distance_elev_diff.csv")
```

```{r}
pop_sum_1 <- pop_sum_1 %>%
  left_join (distance %>% select (distance_km, Pop), by = "Pop") #join distance with population level data

rep_sum_1 <- rep_sum_1 %>%
  left_join(distance %>% select (distance_km, Pop), by = "Pop") #join distance with replicate level data
```

```{r}
glm_distance <- glm( #fit a binomial generalized linear model
  cbind(events, trials - events) ~ distance_km, #model viable and non-viable seeds count as a function of distance
  family = binomial, #use binomial distribution
  data = rep_sum_1 #use replicate level data (because the data has events and trials but the population level data does not)
)

prediction_distance <- data.frame( #create a prediction dataset for geographic distance
  distance_km = seq(
    min(rep_sum_1$distance_km), #start from the shortest to longest geographic distance
    max(rep_sum_1$distance_km),
    length.out = 100 #generate 100 equally spaced distance values
  )
)

prediction_distance$predicted <- predict( #calculate predicted viability across geographic distance
  glm_distance, #use the fitted binomial glm
  newdata = prediction_distance, #predict viability for the new distance values
  type = "response" #return predicted proportions
)

print(summary(glm_distance))
```
###Geographic distance from the common garden significantly affected seed viability after one year of burial (binomial GLM, P < 0.001).

```{r}
anova(glm_distance, test = "Chisq")
```
###Geographic distance from the common garden significantly affected seed viability following one year of burial (Analysis of Deviance, χ² = 14.13, df = 1, P < 0.001).

```{r}
figuer_distance <- ggplot ( #create a plot of population mean viability against geographic distance
  pop_sum_1, aes (x = distance_km, y = mean_proportion)
) + 
  geom_point(size = 2.5) + #plot the mean proportion of viable seeds for each population
  geom_errorbar( #add SE bar
    aes(
      ymin= mean_proportion - SE,
      ymax = mean_proportion + SE
    ),
    width = 5
  ) +
  geom_text (aes (label = Pop), vjust = -6.1, size = 3) + #label each point with its population name
  geom_line(
  data = prediction_distance,
  aes(
    x = distance_km,
    y = predicted
  ),
  linewidth = 1,
  color = "red" #add the fitted binomial glm prediction line
) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(limits = c(.35, .85)) +
  theme(text=element_text(size=10), axis.text.x = element_text(angle = 45,  hjust = 1)) +
  theme_classic() + #use classic ggplot theme
  labs ( #add labels for x-axis and y-axis
    x = "Geographic Distance from CP2 (km)",
    y = "Proportion Viable"
  )
print(figuer_distance)
```

```{r}

```

