---
title: "CP2_vs_Control_viability"
output: html_document
date: "2026-08-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
CP2_Y1 <- read.csv("../output/CP2_Y1_w_elev.csv") %>%
  filter(!(Pop == "WL1" & Rep == 14)) #filter WL1 rep14, because there is no seeds in WL1 rep14 but we have marked it in the spreadsheet; without filter, the NA will be include in the SE and cause all the reps of WL1 become NA
Control_Y1 <- read.csv("../output/Control_Y1_w_elev.csv")
```

```{r}
control_1 <- Control_Y1 %>%
  mutate (
    viable = Updated_viability_cat == 1, #set only consider category 1 as viable
    Treatment = "Control"
    )

CP2_Y1 <- CP2_Y1 %>%
  mutate(
    viable = Updated_viability_cat == 1,
    Treatment = "CP2_Y1"
  )

compare_1 <- bind_rows (control_1, CP2_Y1)
```

```{r}
compare_sum_1 <- compare_1 %>%
  group_by (Pop, Elev_m, Treatment) %>% #group seeds data by population, elevation and replicates
  summarise (
    trials = n(),
    events = sum (viable, na.rm = TRUE), #count how many seeds are considered as viable (category 1)
    proportion = events/trials, 
    SE = sd(viable, na.rm = TRUE)/sqrt(n()),
    .groups = "drop"
  )
print(compare_sum_1)
```

```{r}
compare_glm_1 <- glm( #fit a binomial generalized linear model
  viable ~ Treatment, #overall change between control and CP2
  family = binomial, 
  data = compare_1
)
print(summary(compare_glm_1)) 
```
###Seeds buried at the common garden had significantly higher viability than laboratory-stored control seeds after one year (binomial GLM, P < 0.001).

```{r}
compare_glm_2 <- glm( #fit a binomial generalized linear model
  viable ~ Treatment * Pop, #test whether effect of burial differ among populations
  family = binomial, 
  data = compare_1
)
print(summary(compare_glm_2)) 
```
###Different source populations shows different viability following one year of burial at the common garden in both CP2 and control group


```{r}
anova(compare_glm_2, test = "Chisq")
```
###Overall seed viability differed significantly between laboratory-stored control seeds and seeds buried at CP2 for one year (binomial GLM, χ² = 38.95, P < 0.001).

###Seed viability differed significantly among source populations (χ² = 91.47, P < 0.001).

###The Treatment × Population interaction was not statistically significant (χ² = 16.70, P = 0.081), the overall effect of burial was generally consistent across populations.

```{r}
compare_figure_1 <- ggplot(compare_sum_1, aes (x = reorder(Pop, Elev_m), y = proportion, color = Treatment, group = Treatment)) + #create a ggplot use population level summary data with x-axis in elevation by meter and y-axis by mean proportion of viable
  geom_point(position = position_dodge(width = 0.5), size = 2) + #each populations' mean proportion of viable
  geom_errorbar( #add SE bar
    aes(
      ymin= proportion - SE,
      ymax = proportion + SE
    ),
    width = 0.4, position = position_dodge(width = 0.5)
  ) +
  #coord_flip() +
  #geom_line(position = position_dodge(width = 0.4), linewidth = 0.8)+
  theme_classic() + #use classic ggplot theme
  labs ( #add labels for x-axis and y-axis
    x = "Population (ordered by ascending elevation in meter)",
    y = "Proportion Viable",
    color = "Treatment"
  )
print(compare_figure_1)
```


