---
title: "Clim_dist_calcs_CrossesGardens"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code for calcuLatitudeing climate distance
Update for 2025 water year 

## Load Libraries
```{r}
library(tidyverse)
library(boot) #for bootstrapping 
library(broom) #for tidy function
library(geosphere) #for calcuLatitudeing geographic distance
library(lmerTest) #for mixed effect models 

set.seed(112) #to get consistent bootstrap results
```

## Load Climate Data
```{r}
all_clim <- read_csv("../output/WtrYr_Clim_2025NoOct.csv")
```

## CP2 Garden Climate
```{r}
#CP2 30 years average (recent)
cp2_recent_Garden <- all_clim %>% 
  filter(pop=="CP2_garden", timeframe == "recent", year >= 1995, year <= 2025, Season == "Water Year") %>% #filter CP2 garden and set the time for recent climate data (1995-2025)
  summarise(#summarise and remove NA value for both temp and ppt seasonality data 
    temp_seasonality_Garden = mean(temp_seasonality, na.rm = TRUE),  
    ppt_seasonality_Garden = mean(ppt_seasonality, na.rm = TRUE)
  )

#CP2 30 years average (historic)
cp2_historic_Garden <- all_clim %>%
  filter (pop == "CP2_garden", timeframe == "historic", year >= 1965, year <= 1994, Season == "Water Year") %>%
  summarise (
    temp_seasonality_Garden = mean (temp_seasonality, na.rm = TRUE),
    ppt_seasonality_Garden = mean (ppt_seasonality, na.rm = TRUE)
  )

cp2_recent_range_prep <- cp2_recent_Garden %>%
  rename_with(~str_remove(.,"_Garden"), everything())
cp2_historic_range_prep <- cp2_historic_Garden %>%
  rename_with(~str_remove(.,"_Garden"), everything())
```

## Gower's Climate Distance

(1/P) \* SUM ((absolute value(Ai - Bi)) / range(i)) for each variable

-   P = number of environmental variables = 13 (without CWD)

-   Ai = 30 year avg of that variable for the home site

-   Bi = July 2023-Dec 2023 avg of that variable for the WL2 garden

-   Range(i) = maximum - minimum of that variable in the whole data set
    (across sites)
    
### Prep the dataframes
```{r}
#pop info 
pops <- all_clim %>% 
  filter(pop != "CP2_garden") %>%
  select(pop, Elev_class, Elev_m, Latitude, Longitude) %>% 
  distinct() %>%
  arrange (pop)
pop_order <- pops$pop #make sure the pop names are correctly link for the futher analysis

#home climates
recent_wtryr_clim <- all_clim %>% 
  filter(timeframe=="recent", year >=1995, year <= 2025, #30 year averages 
         Season=="Water Year", pop != "CP2_garden")
recent_wtryr_clim_nest <- recent_wtryr_clim %>% nest(.by=year) #nest to prepare for bootstrapping 

historic_wtryr_clim <- all_clim %>% 
  filter(timeframe=="historic", year >= 1965, year<=1994,
         Season=="Water Year", pop != "CP2_garden")
historic_wtryr_clim_nest <- historic_wtryr_clim %>% nest(.by=year) #nest to prepare for bootstrapping 
```

### Create the gower_calc function 

```{r}
gowers_calc <- function(data, indices, P, garden_clim, garden_range) { #function with all of the code necessary for calcuLatitudeing gowers distance 
  #data = _clim_nest (recent or historical) - needs to be nested by year; P = # climate variables;
  #garden_clim = garden clim with _WL2 or _UCD at end of column names
  #garden_range = garden data with WL2 or UCD removed from the column names 
  
  data <-data[indices,] # subset per bootstrap indices
  
  data <- data %>% unnest(data) #unnest so the function can access the climate data
  
  data_means <- data %>% 
    group_by(pop, Elev_class, Elev_m, Latitude, Longitude, timeframe) %>% 
    summarise(
      temp_seasonality = mean(temp_seasonality, na.rm = TRUE),
      ppt_seasonality = mean(ppt_seasonality, na.rm = TRUE),
      .groups = "drop"
    )#get 30 year averages for each climate variable 
  
  range_merge <- bind_rows(data_means, garden_range)
  
  garden_home_climate_ranges <- range_merge %>% #calculate ranges
    ungroup() %>% 
  summarise(
    temp_seasonality_range = max(temp_seasonality, na.rm = TRUE) - 
      min (temp_seasonality, na.rm = TRUE),
    ppt_seasonality_range = max(ppt_seasonality, na.rm = TRUE) - 
      min (ppt_seasonality, na.rm = TRUE)
  )
  
  garden_home_climate <- bind_cols(garden_clim, data_means) #add garden climate data to home climate data 
  
  garden_home_climate_with_ranges <- bind_cols(garden_home_climate, garden_home_climate_ranges) #add in ranges 
  
  gowers_calc_each_var <- garden_home_climate_with_ranges %>% #variable by variable calc
    mutate(temp_seasonality_gowers = abs(temp_seasonality_Garden - temp_seasonality) / temp_seasonality_range,
           ppt_seasonality_gowers = abs (ppt_seasonality_Garden - ppt_seasonality) / ppt_seasonality_range) %>% 
  dplyr::select(pop, Elev_class, Elev_m, ends_with("_gowers"))

 gowers_calc_per_pop <- gowers_calc_each_var %>% #final gowers calc 
  mutate(Gowers_Dist=(1/P)*(temp_seasonality_gowers + ppt_seasonality_gowers),
         Gowers_Dist_temp = temp_seasonality_gowers,
         Gowers_Dist_ppt = ppt_seasonality_gowers) %>%
   dplyr::select(pop, Elev_class, Elev_m, Gowers_Dist)
  
 gowers_calc_per_pop %>% pull(Gowers_Dist) #make the result a vector 
}
```


### Bootstrapping 

#### CP2 Garden
```{r}
# water year - recent
gowers.boot_cp2_recent <- boot(
  data=recent_wtryr_clim_nest, 
  statistic=gowers_calc, 
  R=1000, P=2, 
  garden_clim=cp2_recent_Garden,
  garden_range=cp2_recent_range_prep) #will sample each row (year) with replacement 

for(i in 1:11) {
  plot(gowers.boot_cp2_recent, index=i) #distributions look normal for the most part 
}

boot_recent_results_cp2 <- tidy(
  gowers.boot_cp2_recent,
  conf.int=TRUE,
  conf.method="norm"
) %>%  #all pops
  rename(Gowers_Dist = statistic) %>% 
  mutate(timeframe="Recent", Season="Water Year")
boot_gowers_recent_cp2 <- bind_cols(pops, boot_recent_results_cp2) %>% 
  select(pop:Longitude, timeframe, Season, Gowers_Dist, conf.low, conf.high) %>% 
  arrange(Gowers_Dist)

# water year - historic
gowers.boot_cp2_historic <- boot(
  data=historic_wtryr_clim_nest,
  statistic=gowers_calc,
  R=1000, P=2,
  garden_clim=cp2_historic_Garden,
  garden_range=cp2_historic_range_prep) 
#will sample each row (year) with replacement 

for(i in 1:11) {
  plot(gowers.boot_cp2_historic, index=i) #distributions look normal for the most part 
}

boot_historic_results_cp2 <- tidy(
  gowers.boot_cp2_historic,conf.int=TRUE,conf.method="norm"
) %>%  #all pops
  rename(Gowers_Dist = statistic) %>% 
  mutate(timeframe="Historic", Season="Water Year")
boot_gowers_historic_cp2 <- bind_cols(pops, boot_historic_results_cp2) %>% 
  select(pop:Longitude, timeframe, Season, Gowers_Dist, conf.low, conf.high) %>% 
  arrange(Gowers_Dist)
all_CP2_gowers <- bind_rows(
  boot_gowers_recent_cp2, boot_gowers_historic_cp2
)
write_csv(all_CP2_gowers, "../output/CP2_Climate_Distance.csv")
```

### Plot the results 
```{r}
# CP2 gower climate distance
all_CP2_gowers %>% 
  ggplot(aes(x=fct_reorder(pop, Elev_m), y=Gowers_Dist, group=pop, fill=Elev_m)) +
  geom_col(width = 0.7,position = position_dodge(0.75)) +
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high),
                width=.5, position = position_dodge(0.75)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_gradient(low = "#F5A540", high = "#0043F0") +
  labs(y="Gower Climate Distance from CP2", fill="Elevation (m)", x="Population") +
  theme_classic() +
  theme(text=element_text(size=18), axis.text.x = element_text(size= 9, angle = 45,  hjust = 1)) +
  facet_wrap(~timeframe)
#ggsave("../output/WtrYr_Clim_CP2.png", width = 24, height = 8, units = "in")
```

