---
title: "Data_prep_update_viability_category"
output: html_document
date: "2026-08-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
Control_Y1_elev <- read_csv("../output/Control_Y1_w_elev.csv")
CP2_Y1_elev <- read_csv("../output/CP2_Y1_w_elev.csv")
```
Adding a new column with the updated viability category
Following the Brassicaceae Group 2 guidelines from the TZ Testing Handbook, if the cotyledons are more than 1/2 stained red, the seed is considered viable
The current Viability_category column scored partially stained seeds 3
If a seed with Viability_category = 3 and More_50_stain = 1, then in a new column called Updated_viability_cat, these seeds should be scored a 1 and all othere seeds should keep the same number as in Viability_category

To-Do:
Need to check if these seeds have a fully stained hypocotyl/radicle as well - DONE confirmed receiving a 1 in the More_50_stain column meant the hypocotyl is fully stained and at least half of the cotyledons are stained
Add Purple_staining and More_50_staining columns to the control Y1 dataset 

```{r}
CP2_Y1_elev <- CP2_Y1_elev %>%
  mutate(Updated_viability_cat = ifelse(Viability_category == 3 & More_50_stain == 1, 1, Viability_category)) #create new col called Updated_viability_cat and fill the rows with a 1 if conditions are met or copy over the values from the Viability_category column if not
#write_csv(CP2_Y1_elev, "../output/CP2_Y1_w_elev.csv")

Control_Y1_elev <- Control_Y1_elev %>%
  mutate(Updated_viability_cat = ifelse(Viability_category == 3 & More_50_stain == 1, 1, Viability_category)) #create new col called Updated_viability_cat and fill the rows with a 1 if conditions are met or copy over the values from the Viability_category column if not
#write_csv(Control_Y1_elev, "../output/Control_Y1_w_elev.csv")
```
