---
title: "Data_prep_climate"
output: html_document
date: "2026-08-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and functions
```{r}
library(tidyverse)
library(raster) #for the CV function and extraction code 

sem <- function(x, na.rm=FALSE) { #standard error function 
  sd(x,na.rm=na.rm)/sqrt(length(na.omit(x)))} 
```

## Load the pop and location data

```{r}
pop_garden_locs_rowids <- read_csv("../input/Sites_coordinate_data.csv") %>% #load in site info for pops and garden
  dplyr::select("Pop", "Elev_m", "Elev_class", "Latitude", "Longitude") %>% #keep only relevant columns
  mutate(Pop = ifelse(Pop == "CP2", "CP2_garden", Pop)) %>% #rename CP2 to CP2 garden
  rowid_to_column() %>% #add row names
  mutate(rowid = as.character(rowid))
```

## Extract WY 2023-2025 from asc files
This is adapted from John's "Flint_Data_Extraction_from_Box.Rmd" code. 
Have to download the data (and unzip it) before running the below code chunks
Data is available on the IntBio Box or online at https://www.sciencebase.gov/catalog/item/69fce794b66b010eddb81d1a 
```{r, eval=FALSE}
locs <- pop_garden_locs_rowids %>% 
  rename(latitude = Latitude, longitude = Longitude) %>% 
  # Make duplicate lat/long columns that will be transformed to match the Flint equal area projection
  mutate(latitude=as.numeric(latitude), 
         longitude=as.numeric(longitude)) %>% 
  mutate(latitude_aea = latitude,
         longitude_aea = longitude)

## define projection of flint data
flint_prj <- '+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs' 

## Designate latitude and longitude columns as coordinates
coordinates(locs) <- ~longitude_aea + latitude_aea

## Define original projection (which is unprojected)
proj4string(locs) <- CRS("+proj=longlat +ellps=WGS84")

## Transform locations from original projection to be consistent with the flint data
locs_t <- spTransform(locs, CRS(flint_prj))
plot(locs_t)

## Make data frame from spatial points object, omit "optional" column
### This data frame has the original coordinates but also transformed coordinates
locs_p <- data.frame(locs_t) %>%
  dplyr::select(-optional)
```

### Extract asc values
```{r, eval=FALSE}
setwd("C:/Users/marik/R/EVE_Scholar_2026/input/Climate_data") #wherever the unzipped Flint files are. This line will have to change if someone else runs the code

climate_variables <- c("aet", "cwd", "pck", "pet", 
                       "ppt", "tmn", "tmx") #if you download additional variables like, rch or run, add those here too

path_name <- c("aet_23_25","cwd_23_25","pck_23_25",
               "pet_23_25","ppt_23_25","tmn_23_25","tmx_23_25") #these should be the folders that are made when you unzip the .zip files, there should be a bunch of .asc files in these folders 

#for testing the code (can ignore this)
#path <- c("/Users/bqc/Library/Mobile Documents/com~apple~CloudDocs/Documents/UCDavis_Postdoc/crosses_common-gardens/input/Flint_WY2025/aet_WY2025") 

# Month named vectors
months <- c("jan" = 1, "feb" = 2, "mar" = 3, "apr" = 4,
            "may" = 5, "jun" = 6, "jul" = 7, "aug" = 8,
            "sep" = 9, "oct" = 10, "nov" = 11, "dec" = 12)

walk(path_name, .progress = TRUE, .f = function(path) {
     
     # Extract variable name from path
     var_name <- str_extract(path, "[a-z]+(?=_23_25)")
     
     # Make list for all files pertaining to climate variable
     file_list <- list.files(
         path = path,                       # CHANGED
         pattern = "\\.asc$",               # CHANGED
         full.names = TRUE                  # CHANGED
     )
     
     # Define a vector for populating pop column
     pops <- rep(locs_p$Pop, times = length(file_list))
     
     # Create empty dataframe
     hist_clim <- data.frame(matrix(
         NA,
         nrow = length(pops),
         ncol = 5
     ))
     
     colnames(hist_clim) <- c("pop", "year", "month", "var", "var.data")
     
     hist_clim$pop <- pops
     hist_clim$var <- var_name
     
     # For loop filling in empty data frame
     for (j in seq_along(file_list)) {       # CHANGED
         
         file_name <- basename(file_list[j])
         
         # CHANGED: no need for dec_file_name
         # dec_file_name <- sub('.gz$', '', file_list[j], fixed = TRUE) #files are no longer gunzipped so read .asc files directly
         
         # CHANGED: create row indices once
         rows <- ((j - 1) * nrow(locs_p) + 1):(j * nrow(locs_p))
         
         hist_clim[rows, "var.data"] <-         # CHANGED
             raster::extract(raster(file_list[j]), locs_t)  # CHANGED
         
         hist_clim[rows, "month"] <-            # CHANGED
             months[[str_sub(file_name, 8, 10)]]
         
         hist_clim[rows, "year"] <-             # CHANGED
             as.numeric(str_sub(file_name, 4, 7))
         
         # file.remove(file_list[j])             # KEEP COMMENTED OUT
     }
     
     # Sort the data frame
     hist_clim <- arrange(hist_clim, year, month)
     
     # Save the data frame
     new_file_name <- paste0("Test_", var_name, "_23_25.csv")
     write.csv(hist_clim, new_file_name, row.names = FALSE)
 })
```

### Merge
```{r, eval=FALSE}
#setwd("C:/Users/marik/R/EVE_Scholar_2026/input/Climate_data") #getting a warning with this. I'm just going to specify the path in the next line
file_list <- list.files(path = "C:/Users/marik/R/EVE_Scholar_2026/input/Climate_data", pattern = "_23_25\\.csv$",full.names = TRUE)
all_climate <- map(file_list, read_csv)
all_climate <- reduce(all_climate, bind_rows) 
head(all_climate)
all_climate <- all_climate %>% pivot_wider(names_from = var, values_from = var.data)
write_csv(all_climate, "../input/Climate_data/Flint_All_WY23_25.csv")
```

## Load flint data for years prior 
Right now I only have the Flint data from WY 2023-2025
Brandie's code uses all of the data, I need to upload the Flint WY data prior to 2023 - use Dimensions_All_1895-2022.csv
```{r}
flint_data_1895_2022 <- read_csv("../input/Dimensions_All_1895-2022.csv") %>%
  filter(pop %in% c("BH", "CP2", "DPR", "IH", "LV2", "SC", "SQ3", "TM2", "WL1", "WL2", "WL3", "YOSE1")) %>%
  mutate(pop = ifelse(pop == "CP2", "CP2_garden", pop)) %>%
  dplyr::select(-c("rch","run","str")) %>%
  mutate(month=month(month, label = TRUE, abbr = TRUE)) %>% #convert from nums to char names 
  mutate(month=as.character(month),
         month=tolower(month)) %>%
  left_join(pop_garden_locs_rowids %>% rename(pop=Pop)) #join  in elev_class column
```

### Merge pre and post 2022 flint data
```{r}
all_climate_pops <- read_csv("../input/Climate_data/Flint_All_WY23_25.csv") %>% 
  left_join(pop_garden_locs_rowids %>% rename(pop=Pop)) %>% #join  in elev_class column
  mutate(month=month(month, label = TRUE, abbr = TRUE)) %>% #convert from nums to char names 
  mutate(month=as.character(month),
         month=tolower(month))
  

all_flint_pops <- bind_rows(flint_data_1895_2022, all_climate_pops)
head(all_flint_pops)
```

## Flint Water Year (Nov-Oct)

```{r}
month_nums <- tibble(month=c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec"),
                     month_nums=c(1:12)) #give calendar months a number 

#merge:
flint_all_year_mosnums <- left_join(all_flint_pops, month_nums) %>% arrange(pop, year, month_nums) 

#define the water year as November-October:
flint_all_year_wtr_yr <- flint_all_year_mosnums %>% 
  mutate(wtr_yr = year + month_nums %in% c(11:12)) %>% 
  mutate(timeframe = case_when(
                                wtr_yr < 1965 ~ "beforehistoric",
                                wtr_yr >= 1965 & wtr_yr < 1995 ~ "historic",
                                wtr_yr >= 1995 & wtr_yr <= 2025 ~ "recent")) %>%
                         mutate(timeframe = as.factor(timeframe), month = as.factor(month)) %>%
                         filter(wtr_yr != 1895) %>% #remove years with less than 12 months of data 
                         mutate(months_wtryr = case_when(
                           month == "nov" ~ 1,
                           month == "dec" ~ 2,
                           month == "jan" ~ 3,
                           month == "feb" ~ 4,
                           month == "mar" ~ 5,
                           month == "apr" ~ 6,
                           month == "may" ~ 7, 
                           month == "jun" ~ 8,
                           month == "jul" ~ 9, 
                           month == "aug" ~ 10,
                           month == "sep" ~ 11,
                           month == "oct" ~ 12 )) %>% 
  dplyr::select(pop:month, timeframe, wtr_yr, months_wtryr, aet:tmx, Elev_m:Longitude)  #organize the columns 
summary(flint_all_year_wtr_yr)

#water year is calculated Nov - October to match census dates at WL2. Could have issues with this, but move forward for now. Can be changed later. 
##Note: we don't currently have Oct for 2025!
```

## Bioclim water year calculations

Referenced this r code: <https://github.com/rspatial/dismo/blob/master/R/biovars.R>
Interested in environmental variation, so bio 4 and 15 but will leave all Brandie ran 

```{r}
# Calculating average temp and diurnal range for each month

bioclim_wtr_yr_prep <- flint_all_year_wtr_yr %>% 
  mutate(tavg = (tmn + tmx)/2, t_diurnal = (tmx-tmn))

#Bio 1, 2, 4, 7, 12, 15
bioclim_wtr_yr_calc <- bioclim_wtr_yr_prep %>% 
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  summarise(ann_tmean=mean(tavg),  #Bio1 - Annual Mean Temperature
            mean_diurnal_range=mean(t_diurnal), #Bio2 - Mean Diurnal Range
            temp_seasonality=sd(tavg), #Bio4 - Temperature Seasonality
            temp_ann_range=(max(tmx))-(min(tmn)), #bio7 - temp annual range
            ann_ppt=sum(ppt), #bio12 - annual precip
            ppt_seasonality=cv(ppt+1)) #bio15 - Precipitation Seasonality (+1 to avoid strange CVs for areas where mean rainfaill is < 1)

# Calculating wettest, driest, warmest, and coldest months 
wettest_month <- bioclim_wtr_yr_prep %>%  #there's one wettest month per pop/year
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  slice_max(ppt)
driest_month <- bioclim_wtr_yr_prep %>%  #there's more than one driest month for some pops and years 
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  slice_min(ppt)
warmest_month <- bioclim_wtr_yr_prep %>% #there's more than one warmest month for 1 pop/year
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  slice_max(tavg)
coldest_month <- bioclim_wtr_yr_prep %>% #there's more than one driest month for some pops and years 
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  slice_min(tavg)

# Bio 8, 9, 18, 19
bio8_wtryr <- wettest_month %>% #bio8 = tmean_wettest_month
  dplyr::select(pop, Elev_class, Elev_m, timeframe, wtr_yr, tmean_wettest_month=tavg)
bio9_wtryr <- driest_month %>% #bio9 = tmean_driest_month
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  summarise(tmean_driest_month=mean(tavg)) #taking the average b/c some years where there are multiple driest months 
bio18_wtryr <- warmest_month %>% #bio18 = ppt_warmest_month
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  summarise(ppt_warmest_month=mean(ppt)) #taking the avg b/c more than one warmest month for one pop/year
bio19_wtryr <- coldest_month %>% #bio19 = ppt_coldest_month
  group_by(pop, Elev_class, Elev_m, timeframe, wtr_yr) %>% 
  summarise(ppt_coldest_month=mean(ppt)) #taking the avg b/c more than one coldest month for one pop/year
#merge:
bio8_9_wtryr <- full_join(bio8_wtryr, bio9_wtryr) 
bio18_19_wtryr <- full_join(bio18_wtryr, bio19_wtryr) 
all_periods_wtryr <- full_join(bio8_9_wtryr, bio18_19_wtryr)

#Merge all bioclims
bioclim_wtryr <- full_join(bioclim_wtr_yr_calc, all_periods_wtryr)
summary(bioclim_wtryr)
```

## Merge Flint and Bioclim
```{r}
#summarize flint by year to merge it with the bioclim data
flint_yrly_avgs_wtr_yr <- flint_all_year_wtr_yr %>% 
  filter(timeframe!="beforehistoric") %>% 
  rename(year=wtr_yr, year_cal=year) %>% 
  group_by(pop, Elev_class, Elev_m, Latitude, Longitude, timeframe, year) %>% 
  summarise_at(c("cwd", "pck", "ppt", "tmn", "tmx"), c(mean), na.rm = TRUE)

#subset bioclim water year
bioclim_wtryr_to_merge <- bioclim_wtryr %>% 
  filter(timeframe!="beforehistoric") %>% 
  rename(year=wtr_yr) 

#merge
wtr_yr_clim <- left_join(flint_yrly_avgs_wtr_yr, bioclim_wtryr_to_merge) %>% 
  mutate(Season="Water Year")
write_csv(wtr_yr_clim, "../output/WtrYr_Clim_2025NoOct.csv")
```


