---
title: "Proportional Viable Seeds: Control vs. CP2 Year 1"
author: "Xiqi Zheng"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width=14, fig.height=8, dpi=300)
library(tidyverse)
library(gridExtra)
library(grid)
library(gtable)
library(flextable)
dir.create("../outputs", showWarnings=FALSE)
```

# Analysis overview

This report compares proportional seed viability between laboratory-stored control seeds and CP2 Year 1 buried seeds.

- Each control seed is one replicate.
- Each CP2 `Population × Rep` is one replicate.
- CP2 Reps 1–10 are included.
- Difference is calculated as `CP2 - Control`.
- Error bars represent ±1 SE.

```{r read-data}
control_raw <- read.csv("../data/STTO Viability Data - Y1_Control_lab_stored_seed_viability_data.csv") %>%
  select(Pop, Viability_category)

cp2_raw <- read.csv("../data/STTO Viability Data - CP2_Y1_Rep1-10_viability_data.csv") %>%
  select(Pop, Rep, Viability_category)

pop_order <- c("TM2","BH","SC","IH","LV2","WL1","DPR","WL2","WL3","YOSE1","SQ3")
max_rep <- 10
```

```{r helper-function, include=FALSE}
make_control_cp2_figure <- function(viable_categories, definition_label, output_stub) {

  control_summary <- control_raw %>%
    mutate(
      Viability_category = as.numeric(Viability_category),
      viable = if_else(Viability_category %in% viable_categories, 1, 0)
    ) %>%
    filter(!is.na(Viability_category)) %>%
    group_by(Pop) %>%
    summarise(
      control_n = n(),
      control_prop = mean(viable),
      control_se = sd(viable) / sqrt(control_n),
      .groups = "drop"
    )

  cp2_rep_summary <- cp2_raw %>%
    mutate(
      Rep = as.integer(Rep),
      Viability_category = as.numeric(Viability_category)
    ) %>%
    filter(Rep %in% 1:max_rep) %>%
    group_by(Pop, Rep) %>%
    summarise(
      trials = sum(!is.na(Viability_category)),
      events = sum(Viability_category %in% viable_categories, na.rm=TRUE),
      .groups = "drop"
    ) %>%
    filter(trials > 0) %>%
    mutate(proportion = events / trials)

  cp2_summary <- cp2_rep_summary %>%
    group_by(Pop) %>%
    summarise(
      cp2_n_reps = n(),
      cp2_prop = mean(proportion),
      cp2_se = sd(proportion) / sqrt(cp2_n_reps),
      .groups = "drop"
    )

  summary_data <- full_join(control_summary, cp2_summary, by="Pop") %>%
    mutate(
      Pop = factor(Pop, levels=pop_order),
      difference = cp2_prop - control_prop,
      control_lower = pmax(0, control_prop - control_se),
      control_upper = pmin(1, control_prop + control_se),
      cp2_lower = pmax(0, cp2_prop - cp2_se),
      cp2_upper = pmin(1, cp2_prop + cp2_se)
    ) %>%
    arrange(Pop)

  table_data <- summary_data %>%
    transmute(
      Population = as.character(Pop),
      Control = sprintf("%.3f (%.3f)", control_prop, control_se),
      `CP2 Year 1` = sprintf("%.3f (%.3f)", cp2_prop, cp2_se),
      `Difference\n(CP2 - Control)` = sprintf("%+.3f", difference)
    )

  table_theme <- ttheme_minimal(
    base_size = 10,
    core = list(
      fg_params = list(hjust=0.5, x=0.5),
      bg_params = list(fill="white", col="grey80")
    ),
    colhead = list(
      fg_params = list(fontface="bold", hjust=0.5, x=0.5),
      bg_params = list(fill="grey95", col="grey65")
    )
  )

  table_plot <- tableGrob(table_data, rows=NULL, theme=table_theme)

  # Population names are left in the table theme's default font.
  # Directly changing gp$fontface can conflict with an existing gp$font
  # setting in some versions of grid/gridExtra.

  difference_cells <- which(table_plot$layout$name == "core-fg" & table_plot$layout$l == 4)
  for (i in seq_along(difference_cells)) {
    value <- summary_data$difference[i]
    if (!is.na(value)) {
      table_plot$grobs[[difference_cells[i]]]$gp$col <- ifelse(value < 0, "firebrick", "blue3")
    }
  }

  plot_data <- summary_data %>%
    select(Pop, control_prop, control_lower, control_upper, cp2_prop, cp2_lower, cp2_upper) %>%
    pivot_longer(
      cols = -Pop,
      names_to = c("Condition", ".value"),
      names_pattern = "(control|cp2)_(prop|lower|upper)"
    ) %>%
    mutate(
      Condition = recode(Condition, control="Control", cp2="CP2 Year 1"),
      Pop = factor(Pop, levels=rev(pop_order))
    )

  forest_plot <- ggplot(plot_data, aes(x=prop, y=Pop, color=Condition, shape=Condition)) +
    geom_vline(xintercept=0.5, linetype="dashed", color="grey75") +
    geom_errorbar(
      aes(xmin=lower, xmax=upper),
      orientation="y",
      width=0.13,
      linewidth=0.6,
      position=position_dodge(width=0.38)
    ) +
    geom_point(size=3.1, position=position_dodge(width=0.38)) +
    scale_color_manual(values=c("Control"="#3B66C4", "CP2 Year 1"="#2F9143")) +
    scale_shape_manual(values=c("Control"=16, "CP2 Year 1"=15)) +
    scale_x_continuous(limits=c(0,1), breaks=seq(0,1,0.2)) +
    labs(x="Proportion viable", y=NULL, color=NULL, shape=NULL) +
    theme_classic(base_size=12) +
    theme(
      legend.position="top",
      legend.justification="center",
      axis.text.y=element_text(size=11),
      axis.title.x=element_text(size=12),
      plot.margin=margin(5,10,5,5)
    )

  # Save the table and plot as separate files
  png(
    paste0("../outputs/", output_stub, "_table.png"),
    width=1800,
    height=1200,
    res=200
  )
  grid.newpage()
  grid.draw(table_plot)
  dev.off()

  ggsave(
    paste0("../outputs/", output_stub, "_plot.png"),
    forest_plot,
    width=8,
    height=6,
    dpi=300
  )

  ggsave(
    paste0("../outputs/", output_stub, "_plot.pdf"),
    forest_plot,
    width=8,
    height=6
  )

  write.csv(
    summary_data,
    paste0("../outputs/", output_stub, "_summary.csv"),
    row.names=FALSE
  )

  list(
    summary_data=summary_data,
    table=table_plot,
    plot=forest_plot,
    definition_label=definition_label
  )
}
```

# Definition 1: Category 1 only

```{r definition-1, include=FALSE}
result_1 <- make_control_cp2_figure(
  viable_categories=c(1),
  definition_label="Viable = Category 1 only",
  output_stub="Control_CP2_Category1"
)
```

## Summary table

```{r definition-1-table, fig.width=8, fig.height=5}
table1 <- result_1$summary_data %>%
  transmute(
    Population = Pop,
    Control = sprintf("%.3f (%.3f)", control_prop, control_se),
    `CP2 Year 1` = sprintf("%.3f (%.3f)", cp2_prop, cp2_se),
    diff = difference,
    `Difference (CP2 - Control)` = sprintf("%+.3f", difference)
  )

flextable(table1) %>%
  color(
    i = ~ diff > 0,
    j = "Difference (CP2 - Control)",
    color = "blue"
  ) %>%
  color(
    i = ~ diff < 0,
    j = "Difference (CP2 - Control)",
    color = "red"
  ) %>%
  delete_columns("diff") %>%
  autofit()
```

## Proportional viability plot

```{r definition-1-plot, fig.width=8, fig.height=6}
result_1$plot +
  labs(
    title="Proportional Viable Seeds: Control vs. CP2 Year 1",
    subtitle=result_1$definition_label
  )
```

*Proportion viable = viable seeds / total tested seeds. Error bars represent ±1 SE. Difference = CP2 − Control.*

# Definition 2: Categories 1 and 2

```{r definition-12, include=FALSE}
result_12 <- make_control_cp2_figure(
  viable_categories=c(1,2),
  definition_label="Viable = Categories 1 and 2",
  output_stub="Control_CP2_Categories1-2"
)
```

## Summary table

```{r definition-12-table, fig.width=8, fig.height=5}
table1 <- result_12$summary_data %>%
  transmute(
    Population = Pop,
    Control = sprintf("%.3f (%.3f)", control_prop, control_se),
    `CP2 Year 1` = sprintf("%.3f (%.3f)", cp2_prop, cp2_se),
    diff = difference,
    `Difference (CP2 - Control)` = sprintf("%+.3f", difference)
  )

flextable(table1) %>%
  color(
    i = ~ diff > 0,
    j = "Difference (CP2 - Control)",
    color = "blue"
  ) %>%
  color(
    i = ~ diff < 0,
    j = "Difference (CP2 - Control)",
    color = "red"
  ) %>%
  delete_columns("diff") %>%
  autofit()
```

## Proportional viability plot

```{r definition-12-plot, fig.width=8, fig.height=6}
result_12$plot +
  labs(
    title="Proportional Viable Seeds: Control vs. CP2 Year 1",
    subtitle=result_12$definition_label
  )
```

*Proportion viable = viable seeds / total tested seeds. Error bars represent ±1 SE. Difference = CP2 − Control.*

# Definition 3: Categories 1, 2, and 3

```{r definition-123, include=FALSE}
result_123 <- make_control_cp2_figure(
  viable_categories=c(1,2,3),
  definition_label="Viable = Categories 1, 2, and 3",
  output_stub="Control_CP2_Categories1-2-3"
)
```

## Summary table

```{r definition-123-table, fig.width=8, fig.height=5}
table1 <- result_123$summary_data %>%
  transmute(
    Population = Pop,
    Control = sprintf("%.3f (%.3f)", control_prop, control_se),
    `CP2 Year 1` = sprintf("%.3f (%.3f)", cp2_prop, cp2_se),
    diff = difference,
    `Difference (CP2 - Control)` = sprintf("%+.3f", difference)
  )

flextable(table1) %>%
  color(
    i = ~ diff > 0,
    j = "Difference (CP2 - Control)",
    color = "blue"
  ) %>%
  color(
    i = ~ diff < 0,
    j = "Difference (CP2 - Control)",
    color = "red"
  ) %>%
  delete_columns("diff") %>%
  autofit()
```

## Proportional viability plot

```{r definition-123-plot, fig.width=8, fig.height=6}
result_123$plot +
  labs(
    title="Proportional Viable Seeds: Control vs. CP2 Year 1",
    subtitle=result_123$definition_label
  )
```

*Proportion viable = viable seeds / total tested seeds. Error bars represent ±1 SE. Difference = CP2 − Control.*

# Output files

The report saves three PNG files, three PDF files, and three CSV summary tables in the `outputs` folder.
