---
title: "Control vs CP2 Year 1 Seed Viability"
author: "Xiqi Zheng"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

library(tidyverse)
library(emmeans)
library(broom)
```

# Analysis overview

This report combines and compares the lab condition stored control seeds and the CP2 Year 1 buried seeds (rep 1-10 for now, 8/3/2026).

- Each control seed is one replicate, with `trials = 1`.
- Each CP2 `Population × Rep` is one replicate, summarized as viable seeds (`events`) out of tested seeds (`trials`).
- Control is the treatment reference.
- CP2 is compared with the corresponding control within each population.

The analysis is repeated using three definitions of viability:

1. Category 1 only
2. Categories 1 and 2
3. Categories 1, 2, and 3

```{r read-data}
control_raw <- read.csv(
  "../data/STTO Viability Data - Y1_Control_lab_stored_seed_viability_data.csv"
) %>%
  select(Pop, Viability_category)

cp2_raw <- read.csv(
  "../data/STTO Viability Data - CP2_Y1_Rep1-10_viability_data.csv"
) %>%
  select(Pop, Rep, Viability_category)

pop_order <- c(
  "TM2", "BH", "SC", "IH", "LV2",
  "WL1", "DPR", "WL2", "WL3", "YOSE1", "SQ3"
)
```

```{r helper-function, include=FALSE}
run_combined_analysis <- function(viable_categories, definition_label) {

  # Control: each seed is one replicate
  control_data <- control_raw %>%
    mutate(
      Viability_category = as.numeric(Viability_category),
      events = if_else(
        Viability_category %in% viable_categories,
        1,
        0
      ),
      trials = 1,
      Treatment = "Control"
    ) %>%
    filter(!is.na(Viability_category)) %>%
    select(Pop, Treatment, events, trials)

  # CP2: each Population × Rep is one replicate
  cp2_data <- cp2_raw %>%
    mutate(
      Rep = as.integer(Rep),
      Viability_category = as.numeric(Viability_category)
    ) %>%
    filter(Rep %in% 1:10) %>%
    group_by(Pop, Rep) %>%
    summarise(
      trials = sum(!is.na(Viability_category)),
      events = sum(
        Viability_category %in% viable_categories,
        na.rm = TRUE
      ),
      .groups = "drop"
    ) %>%
    filter(trials > 0) %>%
    mutate(Treatment = "CP2") %>%
    select(Pop, Treatment, events, trials)

  combined_data <- bind_rows(
    control_data,
    cp2_data
  ) %>%
    mutate(
      Pop = factor(Pop, levels = pop_order),
      Treatment = factor(
        Treatment,
        levels = c("Control", "CP2")
      )
    )

  model <- glm(
    cbind(events, trials - events) ~ Pop * Treatment,
    family = binomial,
    data = combined_data
  )

  estimates <- emmeans(
    model,
    ~ Treatment | Pop,
    type = "response"
  )

  contrasts <- contrast(
    estimates,
    method = "revpairwise",
    adjust = "holm"
  )

  plot_data <- as.data.frame(estimates)

  figure <- ggplot(
    plot_data,
    aes(
      x = Pop,
      y = prob,
      color = Treatment,
      group = Treatment
    )
  ) +
    geom_errorbar(
      aes(
        ymin = asymp.LCL,
        ymax = asymp.UCL
      ),
      width = 0.12,
      position = position_dodge(width = 0.45)
    ) +
    geom_point(
      size = 3,
      position = position_dodge(width = 0.45)
    ) +
    scale_y_continuous(
      limits = c(0, 1),
      breaks = seq(0, 1, 0.25),
      labels = scales::label_percent()
    ) +
    labs(
      title = "Control vs CP2 Year 1 Seed Viability",
      subtitle = definition_label,
      x = "Population",
      y = "Estimated viability",
      color = "Condition"
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold")
    )

  list(
    model = model,
    estimates = estimates,
    contrasts = contrasts,
    figure = figure
  )
}
```

# Definition 1: Category 1 only

```{r definition-1-analysis, include=FALSE}
result_1 <- run_combined_analysis(
  viable_categories = c(1),
  definition_label = "Viable = Category 1 only"
)
```

## Figure

```{r definition-1-figure}
result_1$figure
```

## Overall model test

```{r definition-1-model}
tidy(anova(result_1$model, test = "Chisq")) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  knitr::kable()
```

## Control vs CP2 within each population

```{r definition-1-contrasts}
as.data.frame(result_1$contrasts) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  knitr::kable()
```

# Definition 2: Categories 1 and 2

```{r definition-12-analysis, include=FALSE}
result_12 <- run_combined_analysis(
  viable_categories = c(1, 2),
  definition_label = "Viable = Categories 1 and 2"
)
```

## Figure

```{r definition-12-figure}
result_12$figure
```

## Overall model test

```{r definition-12-model}
tidy(anova(result_12$model, test = "Chisq")) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  knitr::kable()
```

## Control vs CP2 within each population

```{r definition-12-contrasts}
as.data.frame(result_12$contrasts) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  knitr::kable()
```

# Definition 3: Categories 1, 2, and 3

```{r definition-123-analysis, include=FALSE}
result_123 <- run_combined_analysis(
  viable_categories = c(1, 2, 3),
  definition_label = "Viable = Categories 1, 2, and 3"
)
```

## Figure

```{r definition-123-figure}
result_123$figure
```

## Overall model test

```{r definition-123-model}
tidy(anova(result_123$model, test = "Chisq")) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  knitr::kable()
```

## Control vs CP2 within each population

```{r definition-123-contrasts}
as.data.frame(result_123$contrasts) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  knitr::kable()
```

# Notes

- Control is the reference treatment.
- Each contrast compares CP2 with the corresponding control within the same population.
- The `Pop × Treatment` interaction tests whether the Control-to-CP2 change differs among populations.
