---
title: "CP2 Year 1 Seed Viability (Reps 1–10)"
author: "Xiqi Zheng"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(broom)
```

# Analysis overview

Each **Population × Rep** is treated as one replicate.

- Trials = total number of seeds tested
- Events = number of viable seeds

The analysis is repeated using three definitions of viability:

1. Category 1 only
2. Categories 1 and 2
3. Categories 1, 2, and 3

```{r}
seed_data <- read.csv("../data/STTO Viability Data - CP2_Y1_Rep1-10_viability_data.csv")

pop_order <- c("TM2","BH","SC","IH","LV2","WL1","DPR","WL2","WL3","YOSE1","SQ3")

seed_data <- seed_data %>%
  mutate(Pop=factor(Pop,levels=pop_order))
```

```{r}
summarize_rep <- function(data, viable_categories){

  rep_summary <- data %>%
    filter(Rep %in% 1:10) %>%
    group_by(Pop,Rep) %>%
    summarise(
      trials=sum(!is.na(Viability_category)),
      events=sum(Viability_category %in% viable_categories, na.rm=TRUE),
      .groups="drop"
    ) %>%
    filter(trials>0) %>%
    mutate(proportion=events/trials)

  population_summary <- rep_summary %>%
    group_by(Pop) %>%
    summarise(
      Sample_size=n(),
      Mean=mean(proportion),
      SE=sd(proportion)/sqrt(n()),
      Lower_95_CI=pmax(0,Mean-1.96*SE),
      Upper_95_CI=pmin(1,Mean+1.96*SE),
      .groups="drop"
    )

  figure <- ggplot(rep_summary, aes(x = Pop, y = proportion)) +

  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    linewidth = 0.6
  ) +

  stat_boxplot(
    geom = "errorbar",
    width = 0.25,
    linewidth = 0.6
  ) +

  geom_jitter(
    width = 0.12,
    height = 0,
    size = 2,
    alpha = 0.6,
    color = "grey40"
  ) +

  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.25)
  ) +

  theme_classic(base_size = 12) +

  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  ) +

  labs(
    x = "Population",
    y = "Proportion viable"
  )

lm_model <- lm(
  proportion ~ Pop,
  data = rep_summary
)

glm_model <- glm(
  cbind(events, trials - events) ~ Pop,
  family = binomial,
  data = rep_summary
)

list(
  population_summary = population_summary,
  figure = figure,
  lm = lm_model,
  glm = glm_model
)

  lm_model <- lm(proportion~Pop,data=rep_summary)

  glm_model <- glm(
    cbind(events,trials-events)~Pop,
    family=binomial,
    data=rep_summary
  )

  list(
    population_summary=population_summary,
    figure=figure,
    lm=lm_model,
    glm=glm_model
  )
}
```

# Definition 1: Category 1 only

```{r}
result_1 <- summarize_rep(seed_data,1)
```

## Population summary

```{r}
knitr::kable(result_1$population_summary,digits=3)
```

## Figure

```{r}
result_1$figure +
  labs(title="CP2 Year 1 Seed Viability",
       subtitle="Viable = Category 1")
```

## Linear model and ANOVA

```{r}
knitr::kable(tidy(anova(result_1$lm)))
```

## Binomial logistic regression

```{r}
knitr::kable(tidy(anova(result_1$glm,test="Chisq")))
```

# Definition 2: Categories 1 and 2

```{r}
result_12 <- summarize_rep(seed_data,c(1,2))
```

## Population summary

```{r}
knitr::kable(result_12$population_summary,digits=3)
```

## Figure

```{r}
result_12$figure +
  labs(title="CP2 Year 1 Seed Viability",
       subtitle="Viable = Categories 1 and 2")
```

## Linear model and ANOVA

```{r}
knitr::kable(tidy(anova(result_12$lm)))
```

## Binomial logistic regression

```{r}
knitr::kable(tidy(anova(result_12$glm,test="Chisq")))
```

# Definition 3: Categories 1, 2 and 3

```{r}
result_123 <- summarize_rep(seed_data,c(1,2,3))
```

## Population summary

```{r}
knitr::kable(result_123$population_summary,digits=3)
```

## Figure

```{r}
result_123$figure +
  labs(title="CP2 Year 1 Seed Viability",
       subtitle="Viable = Categories 1, 2 and 3")
```

## Linear model and ANOVA

```{r}
knitr::kable(tidy(anova(result_123$lm)))
```

## Binomial logistic regression

```{r}
knitr::kable(tidy(anova(result_123$glm,test="Chisq")))
```
