---
title: "CP2 Year 1 Control Seed Viability"
author: "Xiqi Zheng"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 6, dpi = 300)

library(tidyverse)
library(broom)
library(flextable)
library(knitr)
```

# Analysis overview

Each control seed is treated as one replicate. Seed viability is therefore a binary response: 1 = viable and 0 = not viable.

The analysis is repeated using three definitions of viability:

1. Category 1 only
2. Categories 1 and 2
3. Categories 1, 2, and 3

Because the response is binary, the binomial logistic regression is treated as the primary model. A linear model and ANOVA are also included as requested.

```{r read-data}
control_data <- read.csv(
  "../data/STTO Viability Data - Y1_Control_lab_stored_seed_viability_data.csv"
)

pop_order <- c(
  "TM2", "BH", "SC", "IH", "LV2",
  "WL1", "DPR", "WL2", "WL3", "YOSE1", "SQ3"
)

control_data <- control_data %>%
  mutate(
    Pop = factor(Pop, levels = pop_order),
    Viability_category = as.numeric(Viability_category)
  )
```

```{r analysis-function, include=FALSE}
run_analysis <- function(data, viable_categories) {
  analyzed_data <- data %>%
    mutate(
      viable = case_when(
        is.na(Viability_category) ~ NA_real_,
        Viability_category %in% viable_categories ~ 1,
        TRUE ~ 0
      )
    )

  population_summary <- analyzed_data %>%
    group_by(Pop) %>%
    summarise(
      n = sum(!is.na(viable)),
      proportion_viable = mean(viable, na.rm = TRUE),
      se = sqrt(proportion_viable * (1 - proportion_viable) / n),
      lower_ci = pmax(0, proportion_viable - 1.96 * se),
      upper_ci = pmin(1, proportion_viable + 1.96 * se),
      .groups = "drop"
    ) %>%
    mutate(
      `Viability (%)` = round(proportion_viable * 100, 1),
      `Lower 95% CI (%)` = round(lower_ci * 100, 1),
      `Upper 95% CI (%)` = round(upper_ci * 100, 1)
    )

  figure <- ggplot(analyzed_data, aes(x = Pop, y = viable)) +
    geom_jitter(width = 0.15, height = 0.04, alpha = 0.20,
                size = 1.4, na.rm = TRUE) +
    geom_errorbar(
      data = population_summary,
      aes(x = Pop, ymin = lower_ci, ymax = upper_ci),
      width = 0.15,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = population_summary,
      aes(x = Pop, y = proportion_viable),
      size = 3,
      inherit.aes = FALSE
    ) +
    scale_y_continuous(
      limits = c(-0.08, 1.08),
      breaks = seq(0, 1, 0.25)
    ) +
    labs(
      x = "Population (ordered by elevation)",
      y = "Proportion viable"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  linear_model <- lm(viable ~ Pop, data = analyzed_data)
  linear_anova <- anova(linear_model)

  logistic_model <- glm(
    viable ~ Pop,
    data = analyzed_data,
    family = binomial
  )

  logistic_test <- anova(logistic_model, test = "Chisq")

  list(
    summary = population_summary,
    figure = figure,
    lm = linear_model,
    lm_anova = linear_anova,
    glm = logistic_model,
    glm_test = logistic_test
  )
}

format_summary_table <- function(result) {
  result$summary %>%
    select(
      Population = Pop,
      `Sample size` = n,
      `Viability (%)`,
      `Lower 95% CI (%)`,
      `Upper 95% CI (%)`
    ) %>%
    flextable() %>%
    theme_booktabs() %>%
    autofit()
}

format_lm_table <- function(result) {
  tidy(result$lm_anova) %>%
    mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
    flextable() %>%
    theme_booktabs() %>%
    autofit()
}

format_glm_test <- function(result) {
  tidy(result$glm_test) %>%
    mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
    flextable() %>%
    theme_booktabs() %>%
    autofit()
}

format_glm_coefficients <- function(result) {
  tidy(result$glm) %>%
    mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
    flextable() %>%
    theme_booktabs() %>%
    autofit()
}
```

# Definition 1: Category 1 only

```{r category-1-analysis, include=FALSE}
result_1 <- run_analysis(control_data, 1)
```

## Population summary

```{r category-1-table}
format_summary_table(result_1)
```

## Figure

The small jittered points represent individual seeds. The larger points show the mean viability proportion for each population, and the vertical bars show approximate 95% confidence intervals.

```{r category-1-figure, fig.cap="Control seed viability when only Category 1 is considered viable."}
result_1$figure +
  labs(
    title = "CP2 Year 1 Control Seed Viability",
    subtitle = "Viable = Category 1 only"
  )
```

## Linear model and ANOVA

```{r category-1-lm}
format_lm_table(result_1)
```

## Binomial logistic regression

```{r category-1-glm-test}
format_glm_test(result_1)
```

### Logistic regression coefficients

Coefficients are comparisons with TM2, the reference population.

```{r category-1-glm-coefficients}
format_glm_coefficients(result_1)
```

# Definition 2: Categories 1 and 2

```{r category-12-analysis, include=FALSE}
result_12 <- run_analysis(control_data, c(1, 2))
```

## Population summary

```{r category-12-table}
format_summary_table(result_12)
```

## Figure

```{r category-12-figure, fig.cap="Control seed viability when Categories 1 and 2 are considered viable."}
result_12$figure +
  labs(
    title = "CP2 Year 1 Control Seed Viability",
    subtitle = "Viable = Categories 1 and 2"
  )
```

## Linear model and ANOVA

```{r category-12-lm}
format_lm_table(result_12)
```

## Binomial logistic regression

```{r category-12-glm-test}
format_glm_test(result_12)
```

### Logistic regression coefficients

```{r category-12-glm-coefficients}
format_glm_coefficients(result_12)
```

# Definition 3: Categories 1, 2, and 3

```{r category-123-analysis, include=FALSE}
result_123 <- run_analysis(control_data, c(1, 2, 3))
```

## Population summary

```{r category-123-table}
format_summary_table(result_123)
```

## Figure

```{r category-123-figure, fig.cap="Control seed viability when Categories 1, 2, and 3 are considered viable."}
result_123$figure +
  labs(
    title = "CP2 Year 1 Control Seed Viability",
    subtitle = "Viable = Categories 1, 2, and 3"
  )
```

## Linear model and ANOVA

```{r category-123-lm}
format_lm_table(result_123)
```

## Binomial logistic regression

```{r category-123-glm-test}
format_glm_test(result_123)
```

### Logistic regression coefficients

```{r category-123-glm-coefficients}
format_glm_coefficients(result_123)
```

# Notes

- TM2 is the reference population in the coefficient tables.
- The overall ANOVA and likelihood-ratio tests evaluate whether viability differs among populations.
- SQ3 has a smaller sample size than the other populations, so its estimate is less precise.
- The confidence intervals shown in the figures are descriptive normal-approximation intervals.
- The binomial logistic regression is the preferred inferential analysis for the binary seed-level response.
